\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}

\title{On-Policy Distillation and Memory-Augmented Architectures\\for Continual Learning: A Literature Survey}
\author{Internbot Research Assistant}
\date{March 3, 2026}

\begin{document}
\maketitle

\begin{abstract}
Continual learning (CL) aims to enable models to learn sequentially from non-stationary data streams without catastrophic forgetting of previously acquired knowledge.
Two complementary research directions have emerged as particularly promising: (1)~knowledge distillation-based methods that transfer cumulative knowledge across task boundaries, and (2)~memory-augmented architectures that explicitly store and retrieve past experiences.
This report surveys recent advances in both areas, covering dense distillation strategies, on-policy knowledge transfer, episodic memory management, and the emerging paradigm of modular memory for foundation model agents.
We highlight key findings from eight representative papers and discuss open challenges and future directions.
\end{abstract}

\section{Introduction}

Catastrophic forgetting---the tendency of neural networks to lose previously learned knowledge when trained on new tasks---remains a central obstacle in continual learning~\citep{mccloskey1989catastrophic}.
The problem is especially acute in real-world deployment scenarios where models must adapt to evolving data distributions while retaining prior capabilities.

Two broad families of solutions have gained traction.
\textbf{Distillation-based methods} use a teacher--student framework to regularize learning: the model's outputs on old-task data (or synthetic proxies) are distilled into the updated model, preserving knowledge without storing raw data.
\textbf{Memory-based methods} maintain an explicit buffer of past samples (episodic memory) or learned representations to replay during new-task training.

This report surveys recent work along both axes, with a focus on on-policy distillation techniques and memory-layer architectures.
Section~\ref{sec:distillation} covers distillation approaches, Section~\ref{sec:memory} discusses memory-based methods, Section~\ref{sec:modular} presents the emerging modular-memory paradigm, and Section~\ref{sec:discussion} synthesizes key themes and open questions.

\section{Knowledge Distillation for Continual Learning}
\label{sec:distillation}

Knowledge distillation~\citep{hinton2015distilling} was originally proposed for model compression but has become a cornerstone of continual learning.
The core idea is simple: when learning task $t+1$, the model (student) is regularized to match the outputs of a snapshot of itself from task $t$ (teacher) on representative inputs.

\subsection{Densely Distilling Cumulative Knowledge}

\citet{shi2024densely} identify a key limitation of standard distillation-based CL methods: they typically only distill knowledge from the immediately preceding task model, creating a ``single-hop'' bottleneck.
As the number of tasks grows, knowledge from early tasks is progressively diluted through successive distillation steps.

Their proposed method, \textbf{Densely Distilling Cumulative Knowledge (DDCK)}, addresses this by maintaining distillation connections to multiple prior task models rather than just the most recent one.
The approach draws inspiration from DenseNet-style skip connections, applying them in the temporal (task-sequence) dimension.
Formally, when learning task $t$, the loss includes distillation terms against models from tasks $1, 2, \ldots, t-1$, weighted to balance recency and cumulative coverage:
\begin{equation}
    \mathcal{L}_{\text{DDCK}} = \mathcal{L}_{\text{CE}}(x_t, y_t) + \sum_{i=1}^{t-1} \lambda_i \, \text{KL}\!\left(f_{\theta_i}(x) \,\|\, f_{\theta_t}(x)\right)
\end{equation}
where $\lambda_i$ controls the contribution of each prior model.
Experiments show consistent improvements over single-hop distillation baselines across CIFAR-100, ImageNet-subset, and TinyImageNet benchmarks.

\subsection{On-Policy Distillation in Continual RL}

In the reinforcement learning setting, \citet{traore2019continual} propose using \textbf{policy distillation} to consolidate knowledge across sequential robot manipulation tasks.
The key distinction from supervised CL is that the distillation must be \emph{on-policy}: the student must generate its own trajectories and align its policy with the teacher's action distribution in those states, rather than replaying stored transitions.

Their Sim2Real framework trains policies in simulation for each task, distills them into a single multi-task policy, and then transfers to physical robots.
The on-policy nature of the distillation is critical because off-policy replay of stored transitions leads to distributional shift in the RL setting.
This work demonstrates that policy distillation can serve as an effective knowledge consolidation mechanism for continual RL agents.

\subsection{Federated Continual Learning via Distillation}

\citet{usmanova2022federated} extend distillation-based CL to the \textbf{federated learning} setting, where multiple edge devices learn continually from local data streams.
Each device uses knowledge distillation to consolidate local knowledge, and the distilled models are aggregated at a central server.
This avoids sharing raw data (preserving privacy) while mitigating forgetting across both the temporal (task sequence) and spatial (device) dimensions.

\section{Memory-Based Approaches}
\label{sec:memory}

\subsection{Carousel Memory}

\citet{lee2021carousel} address a fundamental constraint of episodic memory (EM) systems: bounded buffer size.
In standard EM-based CL, when the buffer is full, evicted samples are permanently lost.
\textbf{Carousel Memory (CarM)} introduces a hierarchical memory management strategy that exploits the storage hierarchy of edge devices.

The key insight is that mobile and IoT devices have both fast RAM (small capacity) and slower internal storage (large capacity).
CarM maintains a small active buffer in RAM for training and a larger archive in storage.
Samples are efficiently migrated between the two tiers using a carousel scheduling policy that ensures diverse replay without being bottlenecked by slow storage access.
Experiments across seven CL methods show that CarM consistently improves accuracy by large margins, demonstrating that \emph{how} memory is managed matters as much as \emph{what} is stored.

\subsection{Meta-Learning Representations for Memory Efficiency}

\citet{javed2019meta} take a complementary approach: rather than optimizing the memory buffer, they optimize the \textbf{learned representation} to be amenable to continual learning.
Using meta-learning (specifically, an OML-based objective), they learn feature extractors that produce representations which are both transferable to new tasks and resistant to forgetting.
This reduces the reliance on large episodic memories because the representation itself encodes inductive biases for continual learning.

\subsection{Routing Networks for Task-Specific Memory}

\citet{collier2020routing} use \textbf{routing networks with co-training} to allocate different subnetworks to different tasks.
This implicitly creates task-specific ``memory'' within the network's parameters: knowledge for each task is stored in its assigned routing path, reducing interference between dissimilar tasks.
The co-training mechanism ensures that shared components remain useful across tasks while specialized paths preserve task-specific knowledge.

\section{Modular Memory for Foundation Model Agents}
\label{sec:modular}

The most recent and arguably most forward-looking contribution is the position paper by \citet{dorovatas2026modular}, stemming from a Dagstuhl seminar on continual learning in the era of foundation models.

\subsection{The IWL--ICL Dichotomy}

The paper argues that foundation models have introduced a fundamental dichotomy between two forms of learning:
\begin{itemize}
    \item \textbf{In-Weight Learning (IWL)}: Traditional parameter updates via gradient descent. Stable but slow; the source of catastrophic forgetting.
    \item \textbf{In-Context Learning (ICL)}: Leveraging the context window to condition on examples without parameter updates. Fast and flexible but transient; limited by context length.
\end{itemize}

The central thesis is that neither IWL nor ICL alone is sufficient for continual learning agents.
Instead, \textbf{modular memory} architectures that combine both are needed.

\subsection{A Framework for Modular Memory}

The proposed framework organizes memory into modules with different temporal scales and access patterns:
\begin{itemize}
    \item \textbf{Working memory}: The context window, enabling rapid in-context adaptation.
    \item \textbf{Episodic memory}: Retrieval-augmented stores of past experiences, bridging the gap between transient context and permanent weights.
    \item \textbf{Semantic memory}: Consolidated knowledge encoded in model parameters via selective fine-tuning or adapter modules.
\end{itemize}

This modular design mirrors the memory systems described in cognitive neuroscience and offers a principled architecture for agents that must operate continuously, accumulate experience, and personalize to user needs.

\subsection{Implications for Agent Design}

The paper outlines a roadmap where continual learning moves from the narrow ``task-incremental'' benchmarks of the past decade toward \textbf{open-ended agent learning}.
Key challenges include: (1)~deciding when and what to consolidate from working memory to long-term storage, (2)~efficient retrieval mechanisms for episodic memory at scale, and (3)~preventing semantic drift during IWL updates.

\section{Discussion}
\label{sec:discussion}

\subsection{Convergence of Distillation and Memory}

A recurring theme across the surveyed works is the \textbf{complementarity of distillation and memory}.
Distillation provides a mechanism for knowledge consolidation---compressing what has been learned into a compact form.
Memory provides the raw material for distillation---the exemplars or experiences from which knowledge can be extracted.
Methods that combine both (e.g., using episodic memory to provide replay data for distillation losses) consistently outperform those using either technique in isolation.

\subsection{From Buffers to Architectures}

The field is evolving from treating memory as an external buffer (a bag of stored samples) toward \textbf{architectural integration} of memory mechanisms.
Carousel Memory's hierarchical storage, routing networks' implicit parameter-space memory, and the modular memory framework all point in this direction.
The trend suggests that future CL systems will embed memory as a first-class architectural component rather than an auxiliary training trick.

\subsection{Open Questions}

Several challenges remain:
\begin{enumerate}
    \item \textbf{Scalability of dense distillation}: Methods like DDCK require maintaining or accessing multiple prior models. How to scale this to hundreds of tasks remains unclear.
    \item \textbf{Memory--compute tradeoffs}: Richer memory systems (hierarchical, retrieval-augmented) introduce latency and compute overhead that may conflict with real-time deployment constraints.
    \item \textbf{Evaluation beyond benchmarks}: Most CL research uses closed-set benchmarks (split CIFAR, split ImageNet). Evaluating modular memory systems for open-ended agent learning requires new protocols.
    \item \textbf{Privacy and memory}: Episodic memory stores raw data, raising privacy concerns in federated or multi-user settings. Distillation offers a partial solution by abstracting knowledge away from raw samples.
\end{enumerate}

\section{Conclusion}

This survey has examined two converging threads in continual learning: knowledge distillation and memory-augmented architectures.
Distillation-based methods have matured from single-hop teacher--student transfer to dense cumulative distillation and on-policy RL variants.
Memory-based methods have evolved from simple replay buffers to hierarchical storage systems and meta-learned representations.
The emerging modular memory paradigm, combining in-weight and in-context learning through structured memory modules, offers a compelling vision for the next generation of continual learning agents.

For practitioners, the key takeaway is that effective continual learning increasingly requires \emph{both} mechanisms: distillation for knowledge consolidation and structured memory for experience preservation and retrieval.

\bibliographystyle{plainnat}
\begin{thebibliography}{10}

\bibitem[Collier et~al.(2020)]{collier2020routing}
Mark Collier, Efi Kokiopoulou, Andrea Gesmundo, and Jesse Berent.
\newblock Routing networks with co-training for continual learning.
\newblock \emph{arXiv preprint arXiv:2009.04381}, 2020.

\bibitem[Dorovatas et~al.(2026)]{dorovatas2026modular}
Vaggelis Dorovatas, Malte Schwerin, Andrew~D. Bagdanov, Lucas Caccia, Antonio Carta, Laurent Charlin, Barbara Hammer, Tyler~L. Hayes, Timm Hess, Christopher Kanan, et~al.
\newblock Modular memory is the key to continual learning agents.
\newblock \emph{arXiv preprint arXiv:2603.01761}, 2026.

\bibitem[Hinton et~al.(2015)]{hinton2015distilling}
Geoffrey Hinton, Oriol Vinyals, and Jeff Dean.
\newblock Distilling the knowledge in a neural network.
\newblock \emph{arXiv preprint arXiv:1503.02531}, 2015.

\bibitem[Javed and White(2019)]{javed2019meta}
Khurram Javed and Martha White.
\newblock Meta-learning representations for continual learning.
\newblock In \emph{NeurIPS}, 2019.

\bibitem[Lee et~al.(2021)]{lee2021carousel}
Soobee Lee, Minindu Weerakoon, Jonghyun Choi, Minjia Zhang, Di~Wang, and Myeongjae Jeon.
\newblock Carousel memory: Rethinking the design of episodic memory for continual learning.
\newblock \emph{arXiv preprint arXiv:2110.07276}, 2021.

\bibitem[McCloskey and Cohen(1989)]{mccloskey1989catastrophic}
Michael McCloskey and Neal~J. Cohen.
\newblock Catastrophic interference in connectionist networks: The sequential learning problem.
\newblock In \emph{Psychology of Learning and Motivation}, volume~24, pages 109--165. Academic Press, 1989.

\bibitem[Shi et~al.(2024)]{shi2024densely}
Zenglin Shi, Pei Liu, Tong Su, Yunpeng Wu, Kuien Liu, Yu~Song, and Meng Wang.
\newblock Densely distilling cumulative knowledge for continual learning.
\newblock \emph{arXiv preprint arXiv:2405.09820}, 2024.

\bibitem[Traor\'e et~al.(2019)]{traore2019continual}
Ren\'e Traor\'e, Hugo Caselles-Dupr\'e, Timoth\'ee Lesort, Te~Sun, Guanghang Cai, Natalia~D\'iaz Rodr\'iguez, and David Filliat.
\newblock Continual reinforcement learning deployed in real-life using policy distillation and sim2real transfer.
\newblock \emph{arXiv preprint arXiv:1906.04452}, 2019.

\bibitem[Usmanova et~al.(2022)]{usmanova2022federated}
Anastasiia Usmanova, Fran\c{c}ois Portet, Philippe Lalanda, and German Vega.
\newblock Federated continual learning through distillation in pervasive computing.
\newblock \emph{arXiv preprint arXiv:2207.08181}, 2022.

\end{thebibliography}

\end{document}
